Analysis of Volunteering in the HRS

### 2019-11-22 analysis/techdoc/latex/FEM/estimation_transition_model.tex
First commit in git, migrated from a subversion repository

Since we have a stock sample from the age 25+ population, each respondent goes 
through an individual-specific series of intervals. Hence, we have an 
unbalanced panel over the age range starting from 25 years old. Denote by 
$j_{i0}$ the first age at which respondent $i$ is observed and $j_{iT_i}$ the 
last age when he is observed. Hence we observe outcomes at ages 
$j_i = j_{i0},\ldots,j_{iT_i}$.

We first start with discrete outcomes which are absorbing states (e.g. disease 
diagnostic, mortality, benefit claiming). Record as $h_{i,j_i,m}=1$ if the 
individual outcome $m$ has occurred as of age $j_i$. We assume the 
individual-specific component of the hazard can be decomposed in a time 
invariant and variant part. The time invariant part is composed of the effect of 
observed characteristics $x_i$ that are constant over the entire life course and 
initial conditions $h_{i,j_0,-m}$ (outcomes other than the outcome 
$m$) that are determined before the first age in which each individual is observed
%\footnote{Section \ref{sec:model_development_transition_model} explains why the $h_{i,j_0,-m}$ terms are included.}. 
The time-varying part is the effect of previously diagnosed outcomes $h_{i,j_i-1,-m}$, 
on the hazard for $m$.\footnote{With some abuse of notation, $j_i-1$ denotes 
the previous age at which the respondent was observed.}  We assume an index of 
the form  $z_{m,j_i} = x_i\beta_m + h_{i,j_i-1,-m} \gamma_m + h_{i,j_0,-m}\psi_m$. Hence, the 
latent component of the hazard is modeled as 
\begin{equation}
h^*_{i,j_i,m}= x_i\beta_m + h_{i,j_i-1,-m} \gamma_m + h_{i,j_0,-m}\psi_m + a_{m,j_i} + \varepsilon_{i,j_i,m},
\label{eqn:transition_hzd_latent}
\end{equation}
\[
m = 1,\ldots,M_0 \mbox{, } j_i = j_{i0},\ldots,j_{i,T_i} \mbox{, } i=1,\ldots,N
\]
The term $\varepsilon_{i,j_i,m}$ is a time-varying shock specific to age $j_i$. 
We assume that this last shock is normally distributed and uncorrelated across 
diseases.  We approximate $a_{m,j_i}$ with an age spline with knots at ages 35, 45,
55, 65, and 75.
This simplification is made 
for computational reasons since the joint estimation with unrestricted age fixed 
effects for each condition would imply a large number of parameters.  The 
absorbing outcome, conditional on being at risk, is defined as 
\[
h_{i,j_i,m} = \max\{I(h^*_{i,j_i,m} > 0), h_{i,j_i-1,m}\}
%\label{eqn:transition_outcome}
\]
The occurrence of mortality censors observation of other 
outcomes in a current year. 
% Mortality is recorded from exit interviews.

@@

We have five other types of outcomes:
\begin{enumerate}
\item First, we have binary outcomes which are not an absorbing state, such as starting smoking. We specify latent 
indices as in (\ref{eqn:transition_hzd_latent}) for these outcomes as well but where 
the lag dependent outcome also appears as a right-hand side variable. This allows for state-dependence.

@@

\item The third type of outcomes we consider are censored outcomes, such as financial wealth. For wealth, 
there are a non-negligible number of observations with zero and negative wealth. For these, we consider 
two part models where the latent variable is specified as in (\ref{eqn:transition_hzd_latent}) but model 
probabilities only when censoring does not occur. In total, we have $M$ outcomes.

\item The fourth type of outcomes are continuous outcomes modeled with ordinary least squares.  For example,
we model transitions in log(BMI).  We allow for state-dependence by including the lagged outcome on the right-hand side.

\item The final type of models are categorical, but without an ordering.  For example, an individual can transition
to being out of the labor force, unemployed, or working (either full- or part-time).  In situations like this, we utilize
a multinomial logit model, including the lagged outcome on the right-hand side.

\end{enumerate}

@@

\subsubsection{Further Details on Specific Transition Models}
This section describes the modeling strategy for particular outcomes.

\paragraph{Employment Status}
Ultimately, we wish to simulate if an individual is out of the labor force, unemployed, working part-time, or working full-time at 
time $t$.  We treat the estimation of this as a two-stage process.  In the first stage, we predict if the individual is out of 
the labor force, unemployed, or working for pay using a multinomial logit model.  Then, conditional on working for pay, we estimate if 
the individual is working part- or full-time using a probit model.

\paragraph{Earnings}
We estimate last calendar year earnings models based on the current employment status, controlling for the prior employment status.  
Of particular concern are individuals with no earnings, representing approximately twenty-five percent of the unemployed and 
seventy-eight percent of those out of the labor force.  This group is less than 0.5\% of the full- and part-time populations. We 
use a two-stage process for those out of the labor force and unemployed.  The first stage is a probit that estimates if the 
individual has any earnings.  The second stage is an OLS model of log(earnings) for those with non-zero earnings.  For 
those working full- or part-time, we estimate OLS models of log(earnings).

\paragraph{Relationship Status}
We are interested in three relationship statuses: single, cohabitating, and married.  In each case, we treat the transition
from time $t$ to time $t+1$ as a two-stage process.  In the first stage, we estimate if the individual will remain in their 
current status.  In the second stage, we estimate which of the two other states the individual will transition to, conditional
on leaving their current state.

\paragraph{Childbearing}
We estimate the number of children born in two-years separately for women and men.  We model this using an ordered probit with
three categories: no new births, one birth, and two births.  Based on the PSID data, we found the exclusion of three or more
births in a two-year period to be appropriate.

\subsubsection{Inverse Hyperbolic Sine Transformation}
One problem fitting the wealth distribution is that it has a long right tail and some negative values. We use a 
generalization of the inverse hyperbolic sine transform (IHT) presented in \citet{mackinnon1990transforming}. First denote the variable of 
interest $y$. The hyperbolic sine transform is 
\begin{equation}
y = \sinh(x) = \frac{\exp(x) - \exp(-x)}{2}
\label{eqn:sinh_y}
\end{equation}
The inverse of the hyperbolic sine transform is
\[
x = \sinh^{-1}(y) = h(y) = \log(y + (1+y^2)^{1/2})
\]
Consider the inverse transformation. We can generalize such transformation, first allowing for a 
shape parameter $\theta$,
\begin{equation}
r(y) = h(\theta y)/\theta
\label{eqn:generalized_ihs_shape}
\end{equation}
Such that we can specify the regression model as
\begin{equation}
r(y) = x\beta + \varepsilon, \varepsilon \sim \mathrm{N}(0, \sigma^2)
\label{eqn:ihs_regression_model}
\end{equation}
A further generalization is to introduce a location parameter $\omega$ such that the new 
transformation becomes
\begin{equation}
g(y) = \frac{h(\theta(y+\omega)) - h(\theta\omega)}{\theta h'(\theta \omega)}
\label{eqn:geralized_ihs_loc_scale}
\end{equation}
where $h'(a) = (1+a^2)^{-1/2}$.

@@

% \todo link to transition_estimates.xls on box -- make link unique to the version of the appendix in make process
The included estimates table (estimates\_FAM.xml) gives parameter estimates for the transition models.

>>>>

Since we have a stock sample from the age 51+ population, each respondent goes through an individual-specific series of intervals. Hence, 
we have an unbalanced panel over the age range starting from 51 years old. Denote by $j_{i0}$ the first age at which respondent $i$ is 
observed and $j_{iT_i}$ the last age when he is observed. Hence we observe outcomes at ages $j_i = j_{i0},\ldots,j_{iT_i}$.

We first start with discrete outcomes which are absorbing states (e.g. disease 
diagnostic, mortality, benefit claiming). Record as $h_{i,j_i,m}=1$ if the 
individual outcome $m$ has occurred as of age $j_i$. We assume the 
individual-specific component of the hazard can be decomposed in a time 
invariant and variant part. The time invariant part is composed of the effect of 
observed characteristics $x_i$ that are constant over the entire life course and 
initial conditions $h_{i,j_0,-m}$ (outcomes other than the outcome 
$m$) that are determined before the first age in which each individual is observed
\footnote{Section \ref{sec:model_development_transition_model} explains why the $h_{i,j_0,-m}$ terms are included.}. 
The time-varying part is the effect of previously diagnosed outcomes $h_{i,j_i-1,-m}$, 
on the hazard for $m$.\footnote{With some abuse of notation, $j_i-1$ denotes 
the previous age at which the respondent was observed.}  We assume an index of 
the form  $z_{m,j_i} = x_i\beta_m + h_{i,j_i-1,-m} \gamma_m + h_{i,j_0,-m}\psi_m$. Hence, the 
latent component of the hazard is modeled as 
\begin{equation}
h^*_{i,j_i,m}= x_i\beta_m + h_{i,j_i-1,-m} \gamma_m + h_{i,j_0,-m}\psi_m + a_{m,j_i} + \varepsilon_{i,j_i,m},
\label{eqn:transition_hzd_latent}
\end{equation}
\[
m = 1,\ldots,M_0 \mbox{, } j_i = j_{i0},\ldots,j_{i,T_i} \mbox{, } i=1,\ldots,N
\]
The term $\varepsilon_{i,j_i,m}$ is a time-varying shock specific to age $j_i$. 
We assume that this last shock is normally distributed and uncorrelated across 
diseases.  We approximate $a_{m,j_i}$ with an age spline. After several specification checks, 
knots at age 65 and 75 appear to provide the best fit. This simplification is made 
for computational reasons since the joint estimation with unrestricted age fixed 
effects for each condition would imply a large number of parameters.  The 
absorbing outcome, conditional on being at risk, is defined as 
\[
h_{i,j_i,m} = \max\{I(h^*_{i,j_i,m} > 0), h_{i,j_i-1,m}\}
%\label{eqn:transition_outcome}
\]
The occurrence of mortality censors observation of other 
outcomes in a current year. Mortality is recorded from exit interviews.

@@

We have three other types of outcomes:
\begin{enumerate}
\item First, we have binary outcomes which are not an absorbing state, such as living in a nursing home. We specify latent 
indices as in (\ref{eqn:transition_hzd_latent}) for these outcomes as well but where 
the lag dependent outcome also appears as a right-hand side variable. This allows for state-dependence.

@@

\item The third type of outcomes we consider are censored outcomes, earnings and financial wealth. Earnings are only observed when individuals 
work. For wealth, there are a non-negligible number of observations with zero and negative wealth. For these, we consider two part models 
where the latent variable is specified as in (\ref{eqn:transition_hzd_latent}) but model probabilities only when censoring does not occur. In 
total, we have $M$ outcomes.
\end{enumerate}

@@

\subsubsection{Inverse Hyperbolic Sine Transformation}
One problem fitting the wealth and earnings distribution is that they have a long right tail and wealth has some negative values. We use a 
generalization of the inverse hyperbolic sine transform (IHT) presented in \citet{mackinnon1990transforming}. First denote the variable of 
interest $y$. The hyperbolic sine transform is 
\begin{equation}
y = \sinh(x) = \frac{\exp(x) - \exp(-x)}{2}
\label{eqn:sinh_y}
\end{equation}
The inverse of the hyperbolic sine transform is
\[
x = \sinh^{-1}(y) = h(y) = \log(y + (1+y^2)^{1/2})
\]
Consider the inverse transformation. We can generalize such transformation, first allowing for a 
shape parameter $\theta$,
\begin{equation}
r(y) = h(\theta y)/\theta
\label{eqn:generalized_ihs_shape}
\end{equation}
Such that we can specify the regression model as
\begin{equation}
r(y) = x\beta + \varepsilon, \varepsilon \sim \mathrm{N}(0, \sigma^2)
\label{eqn:ihs_regression_model}
\end{equation}
A further generalization is to introduce a location parameter $\omega$ such that the new 
transformation becomes
\begin{equation}
g(y) = \frac{h(\theta(y+\omega)) - h(\theta\omega)}{\theta h'(\theta \omega)}
\label{eqn:geralized_ihs_loc_scale}
\end{equation}
where $h'(a) = (1+a^2)^{-1/2}$.

@@

% \todo link to transition_estimates.xls on box -- make link unique to the version of the appendix in make process
%The estimates table\footnote{\input{estimates_table_link.tex}} gives parameter estimates for the transition models.

%\newpage
%\input{../tables/FEM/health_transition_est.tex}
%\caption{Coefficient estimates and $t$-statistics for mortality and chronic condition transitions}
%\label{tab:health_transition_est}
%\end{longtable}
%
%\newpage
%\input{../tables/FEM/econ_transition_est.tex}
%\caption{Coefficient estimates and $t$-statistics for economic outcome transitions}
%\label{tab:econ_transition_est}
%\end{longtable}
%
%\newpage
%\input{../tables/FEM/BMI_transition_est.tex}
%\caption{Coefficient estimates and $t$-statistics for BMI outcome transition}
%\label{tab:BMI_transition_est}
%\end{longtable}
%
%\newpage
%\input{../tables/FEM/ordered_transition_est.tex}
%\caption{Coefficient estimates and $t$-statistics for ordered outcome transitions}
%\label{tab:ordered_transition_est}
%\end{longtable}
%
%\newpage
%\input{../tables/FEM/IHS_transition_est.tex}
%\caption{Coefficient estimates and $t$-statistics for transitioning outcomes with shape-preserving Inverse Hyperbolic Sine transformation}
%\label{tab:IHS_transition_est}
%\end{longtable}

### 2019-11-22 analysis/techdoc/latex/FEM/govt_revenue_and_expenditures_medcost_estimation.tex
First commit in git, migrated from a subversion repository

\subsection{Medical costs estimation}
\label{sec:govt_revenue_and_expenditures_medcost_estimation}
In the FAM, a cost module links a person's current state--demographics, economic status, current health, risk factors, and functional status 
to 4 types of individual medical spending. The FAM models: total medical spending (medical spending from all payment sources), Medicare 
spending\footnote{We estimate annual medical spending paid by specific parts of 
Medicare (Parts A, B, and D) and sum to get the total Medicare expenditures.}, 
Medicaid spending (medical spending paid by Medicaid), and out of pocket spending 
(medical spending by the respondent). These estimates are based on pooled weighted 
least squares regressions of each type of spending on risk factors, self-reported 
conditions, and functional status, with spending inflated to constant dollars 
using the medical component of the consumer price index.  We use the 2007-2010 
Medical Expenditure Panel Survey 
% \todo : Change the text to be something like "roughly X respondents per panel, with a new panel starting entering every Y months and lasting Z months between YYYY and YYYY." That way, it is always correct and all we do is change the years of data we're looking at. 
for these regressions for persons 
not Medicare eligible, and the 2007-2010 Medicare Current Beneficiary Survey 
% \todo : Change the text to be something like "roughly X respondents per panel, with a new panel starting entering every Y months and lasting Z months between YYYY and YYYY." That way, it is always correct and all we do is change the years of data we're looking at. 
for spending for those that are eligible for Medicare. Those eligible for 
Medicare include people eligible due to age (65+) or due to disability status. Comparisons of prevalences and question wording across these different sources are provided in Tables \ref{tab:svy_disease_prevalence} and \ref{tab:svy_disease_questions}, respectively.

In the baseline scenario, this spending estimate can be interpreted as the resources consumed by the individual given the manner in which 
medicine is practiced in the United States during the post-part D era (2006-2010). 
Models are estimated for total, Medicaid, out of pocket spending, and for the Medicare spending.

Since Medicare spending has numerous components (Parts A and B are considered here), models are needed to predict enrollment. In 2004, 
98.4\% of all Medicare enrollees, and 99\%+ of aged enrollees, were in Medicare Part A, and thus we assume that all persons eligible for 
Medicare take Part A. We use the 2007-2010 MCBS to model take up of Medicare Part B for both new enrollees into Medicare, as well as 
current enrollees without Part B. Estimates are based on weighted probit regression on various risk factors, demographic, and economic 
conditions. The PSID starting population for the FAM does not contain information on Medicare enrollment. Therefore another model of Part B 
enrollment for all persons eligible for Medicare is estimated via a probit, and used in the first year of simulation to assign initial Part 
B enrollment status. Estimation results are shown in estimates table. The MCBS data over represents the portion enrolled in Part B, having a 97\% 
enrollment rate in 2004 instead of the 93.5\% rate given by Medicare Trustee's Report. In addition to this baseline enrollment probit, we 
apply an elasticity to premiums of -0.10, based on the literature and simulation 
calibration for actual uptake through 2009 \citep{atherly2004effect,buchmueller2006price}. 
The premiums are computed using average Part B costs from the previous time step 
and the means-testing thresholds established by the ACA.

%Since both the MEPS and MCBS are known to under-predict medical spending (see, e.g., \citeauthor{selden2008aligning}, \citeyear{selden2008aligning}, and references therein), we applied adjustment factors to the predicted three types of 
%individual medical spending so that the predicted per-capita spending in FAM equal the corresponding spending in National 
%Health Expenditure Accounts (NHEA) for age group 55-64 in year 2004 and ages 65 and over in year 2010, respectively. Table 
%\ref{tab:NHEA_adjustment} shows how these adjustment factors were determined by using the ratio of expenditures in the NHEA to 
%expenditures predicted in the FAM.

>>>>

\subsection{Medical costs estimation}
\label{sec:govt_revenue_and_expenditures_medcost_estimation}
In the FEM, a cost module links a person's current state--demographics, economic status, current health, risk factors, and functional status 
to 4 types of individual medical spending. The FEM models: total medical spending (medical spending from all payment sources), Medicare 
spending\footnote{We estimate annual medical spending paid by specific parts of 
Medicare (Parts A, B, and D) and sum to get the total Medicare expenditures.}, 
Medicaid spending (medical spending paid by Medicaid), and out of pocket spending 
(medical spending by the respondent). These estimates are based on pooled weighted 
least squares regressions of each type of spending on risk factors, self-reported 
conditions, and functional status, with spending inflated to constant dollars 
using the medical component of the consumer price index.  We use the 2000-2010 
Medical Expenditure Panel Survey 
% \todo : Change the text to be something like "roughly X respondents per panel, with a new panel starting entering every Y months and lasting Z months between YYYY and YYYY." That way, it is always correct and all we do is change the years of data we're looking at. 
for these regressions for persons 
not Medicare eligible, and the 2000-2010 Medicare Current Beneficiary Survey 
% \todo : Change the text to be something like "roughly X respondents per panel, with a new panel starting entering every Y months and lasting Z months between YYYY and YYYY." That way, it is always correct and all we do is change the years of data we're looking at. 
for spending for those that are eligible for Medicare. Those eligible for 
Medicare include people eligible due to age (65+) or due to disability status. Comparisons of prevalences and question wording across these different sources are provided in Tables \ref{tab:svy_disease_prevalence} and \ref{tab:svy_disease_questions}, respectively.

In the baseline scenario, this spending estimate can be interpreted as the resources consumed by the individual given the manner in which 
medicine is practiced in the United States during the post-part D era (2006-2010). 
Models are estimated for total, Medicaid, out of pocket spending, and for the Medicare spending. 
These estimates only use the MCBS dataset.

Since Medicare spending has numerous components (Parts A and B are considered here), models are needed to predict enrollment. In 2004, 
98.4\% of all Medicare enrollees, and 99\%+ of aged enrollees, were in Medicare Part A, and thus we assume that all persons eligible for 
Medicare take Part A. We use the 2007-2010 MCBS to model take up of Medicare Part B for both new enrollees into Medicare, as well as 
current enrollees without Part B. Estimates are based on weighted probit regression on various risk factors, demographic, and economic 
conditions. The HRS starting population for the FEM does not contain information on Medicare enrollment. Therefore another model of Part B 
enrollment for all persons eligible for Medicare is estimated via a probit, and used in the first year of simulation to assign initial Part 
B enrollment status. Estimation results are shown in estimates table. The MCBS data over represents the portion enrolled in Part B, having a 97\% 
enrollment rate in 2004 instead of the 93.5\% rate given by Medicare Trustee's Report. In addition to this baseline enrollment probit, we 
apply an elasticity to premiums of -0.10, based on the literature and simulation 
calibration for actual uptake through 2009 \citep{atherly2004effect,buchmueller2006price}. 
The premiums are computed using average Part B costs from the previous time step 
and the means-testing thresholds established by the ACA.

Since both the MEPS and MCBS are known to under-predict medical spending (see, e.g., \citeauthor{selden2008aligning}, \citeyear{selden2008aligning}, and references therein), we applied adjustment factors to the predicted three types of 
individual medical spending so that the predicted per-capita spending in FEM equal the corresponding spending in National 
Health Expenditure Accounts (NHEA) for age group 55-64 in year 2004 and ages 65 and over in year 2010, respectively. Table 
\ref{tab:NHEA_adjustment} shows how these adjustment factors were determined by using the ratio of expenditures in the NHEA to 
expenditures predicted in the FEM.

### 2019-11-22 analysis/techdoc/latex/FEM/new_cohorts_model_model_and_estimation.tex
First commit in git, migrated from a subversion repository

%To apply trends to the future cohorts, the latent model is written as
%\[
%\mathbf{y}^*_i = \mathbf{\mu} + \mathbf{L}_\Omega \mathbf{\eta}_i.
%\]
%Each marginal has a mean change equal to $\mathrm{E}(\mathbf{y} \mid \mathbf{\mu}) = (1+\tau)g(\mathbf{\mu})$, where $\tau$ is the 
%percent change in the outcome and $g()$ is a non-linear but monotone mapping. Since 
%it is invertible, we can find the vector $\mathbf{\mu}^*$ where $\mathbf{\mu}^* = g^{-1}(\mathrm{E}(y \mid \mu)/(1+\tau))$. We use these new intercepts to simulate new outcomes.

% \todo : add an  example here for clarification

>>>>

To apply trends to the future cohorts, the latent model is written as
\[
\mathbf{y}^*_i = \mathbf{\mu} + \mathbf{L}_\Omega \mathbf{\eta}_i.
\]
Each marginal has a mean change equal to $\mathrm{E}(\mathbf{y} \mid \mathbf{\mu}) = (1+\tau)g(\mathbf{\mu})$, where $\tau$ is the 
percent change in the outcome and $g()$ is a non-linear but monotone mapping. Since 
it is invertible, we can find the vector $\mathbf{\mu}^*$ where $\mathbf{\mu}^* = g^{-1}(\mathrm{E}(y \mid \mu)/(1+\tau))$. We use these new intercepts to simulate new outcomes. 
% \todo : add an  example here for clarification

### 2019-11-22 analysis/techdoc/latex/shared/dynamic_model_background.tex
First commit in git, migrated from a subversion repository

Since then various extensions have been implemented to the original model. The most recent version of the FEM
now projects health outcomes for all Americans aged 51 and older and uses the Health and Retirement 
Study (HRS) as a host dataset rather than the Medicare Current Beneficiary Survey (MCBS).  The work 
has also been extended to include economic outcomes such as earnings, labor force participation and 
pensions. This work was funded by the National Institute on Aging through its support of the RAND 
Roybal Center for Health Policy Simulation (P30AG024968), the Department of Labor through contract 
J-9-P-2-0033, the National Institutes of Aging through the R01 grant ``Integrated Retirement 
Modeling'' (R01AG030824) and the MacArthur Foundation Research Network on an Aging Society.

This document describes the Future Americans Model (FAM), the development of the model to forecast Americans aged 25 
and older.  FAM uses the Panel Survey of Income Dynamics (PSID) as the host dataset.  In addition to modeling
health, health care costs, and economic outcomes, FAM also models life events such as changes in marital status 
and childbearing.  Development of FAM is suported by the National Institutes of Aging through the USC Roybal 
Center for Health Policy Simulation (5P30AG024968-13) and the MacArthur Foundation Research Network on an Aging Society.

% \todo add description of work extending to ages 25+ (FAM) here

>>>>

Since then various extensions have been implemented to the original model. The most recent version 
now projects health outcomes for all Americans aged 51 and older and uses the Health and Retirement 
Study (HRS) as a host dataset rather than the Medicare Current Beneficiary Survey (MCBS).  The work 
has also been extended to include economic outcomes such as earnings, labor force participation and 
pensions. This work was funded by the National Institute on Aging through its support of the RAND 
Roybal Center for Health Policy Simulation (P30AG024968), the Department of Labor through contract 
J-9-P-2-0033, the National Institutes of Aging through the R01 grant ``Integrated Retirement 
Modeling'' (R01AG030824) and the MacArthur Foundation Research Network on an Aging Society. Finally, 
the computer code of the model was transferred from Stata to C++. This report incorporates these new 
development efforts in the description of the model.
% \todo add description of work extending to ages 25+ (FAM) here
